---
title: "Solar Panels"
output: html_document
---

# Setup and test

```{r setup, message=FALSE}
# Setup
library(tidyverse)
library(magrittr)
library(lubridate)
library(ggrepel)
library(ggExtra)
library(naniar)
library(here)
library(readxl)  
```

# Solar Irradiance Cycles by State

## Import and prepare data 
```{r irradiance-import, warning=FALSE}
dni_df <- read_excel(here('data', 'data-solar-irradiance', 'solarsummaries.xlsx'),
                     sheet = 'DNI State',
                     range = 'A1:BO52',
                     col_types = c('text', rep('numeric', 66)))

# Reshape from wide to long
dni_df %<>%
  gather(key = metric, value =  value, -c("State", "State FIPS")) 

# Remove time period from metric name and reshape back to wide-ish
# Now unit of analysis is state-month-estimate instead of state-estimate
dni_df %<>%
  mutate(period = substr(dni_df$metric, 1, str_locate(dni_df$metric, " ") - 1)) %>%
  mutate(attr = substr(dni_df$metric, str_locate(dni_df$metric, " ") + 1, 
                       nchar(dni_df$metric))) %>%
  mutate(period = substr(period, 1, 3)) %>%
  mutate(period = factor(period,
                         levels = c('Jan', 'Feb', 'Mar', 'Apr', 'May', 'Jun', 
                                    'Jul', 'Aug', 'Sep', 'Oct', 'Nov', 'Dec',
                                    'Ann'))) %>%
  select(-metric) %>%
  spread(key = attr, value = value) %>%
  rename(avg = 'Average (kWh/m2/day)',
         max = 'Average Maximum (kWh/m2/day)',
         min = 'Average Minimum (kWh/m2/day)',
         median = 'Average Median (kWh/m2/day)',
         sd = 'Average Standard Deviation (kWh/m2/day)')
```

## Irradiance cycles by state and month

```{r irradiance-plot}
# Plot monthly cycles of direct normal solar irradiance:
# Version with lines colored by intentional highlight
dni_df %>%
  filter(period != "Ann") %>%
  mutate(highlight = ifelse(State %in% c('California', 'Arizona', 'Massachusetts',
                                         'Connecticut', 'New Jersey'),
                            TRUE, FALSE)) %>%
  ggplot() +
  geom_line(aes(x = period, y = avg, group = State, color = highlight),
            show.legend = FALSE,
            na.rm = TRUE) +
  geom_text_repel(data = dni_df[(dni_df$period == 'Dec' & 
                           dni_df$State %in% c('California', 'Arizona', 
                                               'Massachusetts', 'Connecticut', 
                                               'New Jersey')), ],
          aes(label = state.abb[match(State, state.name)], x = Inf, y = avg)) +
  scale_color_manual(values = c('light grey', 'red')) +
  theme_minimal() +
  labs(title = 'States with most solar panels vary widely in solar resource',
       subtitle = 'By average annual Direct Normal Irradiance (DNI), 1998-2009',
       x = 'Month',
       y = 'Average DNI (kWh/m2/day)',
       caption = 'Data source: NREL (US Dept of Energy)')

ggsave(here('output', 'output-intermediate', 'irradiance_by_state_highlight.pdf'), 
       width = 8, height = 6)

# Code to order states by average annual DNI
# dni_df %>% filter(period == 'June') %>% select(c(State, avg)) %>% arrange(-avg)
```

```{r irradiance-plot-2}
# Version with lines colored by region of state
# Highlight states with most panels per 10k pop: CA, AZ, MA, CT, NJ
dni_df %>%
  filter(State != "Alaska") %>%
  filter(period != "Ann") %>%
  mutate(region = state.region[match(State, state.name)]) %>%
  filter(!is.na(region)) %>%
  mutate(highlight = ifelse(State %in% c('California', 'Arizona', 'Massachusetts',
                                         'Connecticut', 'New Jersey'),
                            TRUE, FALSE)) %>%
  mutate(label = ifelse(State %in% c('California', 'Arizona', 'Massachusetts',
                                         'Connecticut', 'New Jersey'),
                        state.abb[match(State, state.name)], "")) %>%
  ggplot(aes(x = period, y = avg, group = State, color = highlight)) +
  geom_line(na.rm = TRUE,
            show.legend = FALSE) +
  geom_text_repel(data = . %>% filter(period == "Dec") ,
                  aes(x = Inf, y = avg, label = label),
                  inherit.aes = FALSE) +
  scale_color_manual(values = c('light grey', 'red')) +
  facet_grid(. ~ region) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = '3 of top 5 states for solar panels receive the least sun',
       subtitle = 'Each line tracks the average direct normal irradiance (DNI) for a specific state.\nStates with the most solar panels per capita are highlighted in red.',
       x = 'Month',
       y = 'Average DNI (kWh/m2/day)',
       caption = 'Data source: NREL (US Dept of Energy)')

# TODO: Get labels for highlighted states only - how to get labels only for data within specific facets?

ggsave(here('output', 'output-intermediate', 
            'irradiance_by_state_region_highlight.pdf'), 
       width = 8, height = 6)
```

```{r irradiance-plot-3}
# Version with lines colored by region, no labels
dni_df %>%
  filter(State != "Alaska", period != "Ann") %>%
  mutate(region = state.region[match(State, state.name)]) %>%
  group_by(State) %>%
  mutate(annual_avg = mean(avg)) %>%
  filter(!is.na(region)) %>%
  ggplot() +
  geom_line(aes(x = period, y = avg, group = State, color = annual_avg),
            na.rm = TRUE) +
  scale_color_gradient(low = '#FFCC00', high = '#FF3300',
                       guide = 'colorbar', 
                       name = 'Avg DNI\n(annual)') +
  facet_grid(. ~ region) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = 'Northeast states receive the least solar radiation overall',
       subtitle = 'By average monthly and annual Direct Normal Irradiance (DNI), 1998-2009',
       x = 'Month',
       y = 'Average DNI (kWh/m2/day)',
       caption = 'Data source: NREL (US Dept of Energy)')

# TODO: Get labels for highlighted states 

ggsave(here('output', 'output-intermediate', 'irradiance_by_state_region.pdf'), 
       width = 8, height = 6)
```


# Solar Panel Installations per 10K people by State, 1969-2016

## Import and prepare data
```{r panels-import, results='hide'}
# Import solar panel data
panels_df <- read_csv(here('data', 'data-solar-panels', 'openpv_all.csv'),
                      col_types = cols_only(state = "c",
                                            date_installed = "c",
                                            zipcode = "c",
                                            cost = "c"),
                      quoted_na = TRUE)
panels_df %<>%
  mutate(date_installed = as.Date(date_installed, format = '%m/%d/%Y')) %>%
  mutate(year_installed = year(date_installed))

head(panels_df)
```

Next, we need to get population data to scale our installation counts by. The metric I'm aiming for is "number of solar panel installations per 100,000 people".

```{r pop-import, results='hide'}
# Prepare population data from non-separated formats
# Source: https://seer.cancer.gov/popdata/popdic.html
pop_df <- read_csv(here('data', 'data-population-estimates',
                     'us.1969_2016.19ages.adjusted.txt'),
                col_types = 'c',
                col_names = FALSE)

pop_df %<>%
  mutate(year = as.numeric(substr(X1, 1, 4)),
         state = substr(X1, 5, 6),
         state_fips = substr(X1, 7, 8),
         county_fips = substr(X1, 9, 11),
         seer_registry = substr(X1, 12, 13),
         race = substr(X1, 14, 14),
         origin = substr(X1, 15, 15),
         sex = substr(X1, 16, 16),
         age = as.numeric(substr(X1, 17, 18)),
         population = as.numeric(substr(X1, 19, 27)))

# Get summary by state-year
pop_by_state <- pop_df %>%
  group_by(year, state) %>%
  summarize(total_pop = sum(population))
```

Joining the new population data to the solar panel data:

```{r panel-join}
# 1. Group by year and fill in empty years with 0 installs
panels_cumul <- panels_df %>%
  group_by(state, year_installed) %>%
  count() %>%
  ungroup() %>%
  complete(state, year_installed, fill = list(n = 0)) %>%
  rename(year = year_installed)
  
# 2. Inner-join population data
# 3a. Get cumulative sum of installations by state
# 3b. Add 1 to each obs to accomodate log scale
panels_cumul %<>%
  inner_join(pop_by_state, by = c('state', 'year')) %>%
  filter(year != 1969) %>%
  mutate(sum_n = ave(n, state, FUN = cumsum)) %>%
  mutate(sum_n_1 = sum_n + 1) %>%
  mutate(panels_per_10k = (sum_n_1 / total_pop) * 10000)
  
# 3. Get top n states by panels per 10,000 population
panels_cumul %>% 
  filter(year == 2016) %>% 
  select(state, panels_per_10k) %>% 
  arrange(-panels_per_10k)
# CA, AZ, MA, CT, NJ
```


## Cumulative total solar panel installations by state and solar resource

```{r panel-plot-1}
# Cumulative totals as of 2016
panels_cumul %>%
  mutate(region = state.region[match(state, state.abb)]) %>%
  filter(!is.na(region)) %>%
  filter(year == 2016) %>%
  ggplot(aes(x = reorder(state, -panels_per_10k), 
             y = panels_per_10k)) +
  geom_bar(stat = 'identity',
           show.legend = FALSE) +
  facet_grid(. ~ region, 
             scales = "free_x", 
             space = "free_x") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(title = 'West and Northeast states well-represented in solar panels',
       subtitle = 'Total solar panel installations per 10,000 people (cumulative as of 2016)',
       x = 'State',
       y = 'Panels per 10,000 people',
       caption = 'Data sources:\n NREL (US Dept of Energy)\nNational Cancer Institute')

ggsave(here('output', 'output-intermediate', 'panels_by_state_region.pdf'), 
       width = 8, height = 6)
```

## Cumulative installations over time by state

```{r panel-plot-2}
# Cumulative installs over time
panels_cumul %>%
  mutate(label = ifelse((panels_cumul$year == 2016 &
                         panels_cumul$state %in% c('CA', 'AZ', 'MA', 'CT', 'NJ')),
                        state, NA)) %>%
  mutate(region = state.region[match(state, state.abb)]) %>%
  filter(!is.na(region)) %>%
  ggplot() +
  geom_line(aes(x = year, y = panels_per_10k, group = state, color = region),
            alpha = 0.5,
            show.legend = FALSE) +
  geom_text_repel(aes(label = label, x = Inf, y = panels_per_10k),
                  na.rm = TRUE) +
  theme_minimal()

# TODO: What is the national average? Which states have above average? 

# panels_cumul %>%
#   filter(year == 2016) %>%
#   select(year, sum_n, total_pop) %>%
#   summarize(total_n = sum(sum_n), 
#             us_pop = sum(total_pop),
#             avg_panels_per_10k = sum(sum_n) / sum(total_pop))
```

```{r panel-plot-3}
# Cumulative installs over time by state facets
panels_cumul %>%
  mutate(region = state.region[match(state, state.abb)]) %>%
  filter(!is.na(region)) %>%
  ggplot() +
  geom_line(aes(x = year, y = panels_per_10k, group = state, color = region),
            show.legend = FALSE) +
  theme_minimal() +
  facet_wrap(. ~ state)

ggsave(here('output', 'output-intermediate', 'panels_by_state_facet.pdf'),
       width = 8, height = 6)
```


# Availability of Personal Tax Credits for Renewable Energy Systems

## Import and prepare data

```{r tax-import, results='hide', warning=FALSE}
policy <- read_csv(here('data', 'data-state-policies', 
                        'correlatesofstatepolicyprojectv2_1.csv')) %>%
  select(c('state', 'year', 
           'corporaterenew_yearadopted',
           'environment_publicbenefit_funds',
           'frenew',
           'frpsexc',
           'frps',
           'netmeter_yearadopted',
           'renewport',
           'rps_yearadopted',
           'personaltax_yearadopted', # Renewable tax credits available?
           'Cons_Renewable'))

```

## Which states provide tax credits for renewable energy systems? And for how long?

```{r tax-plot-1, fig.width=3, fig.height=4}
# Heatmap of policy adoption == True by state/year 
# Allows for non-continuous periods of adoption (e.g. if state stops policy)
policy %>%
  group_by(state) %>%
  filter(year >= 1980) %>%
  select(state, year, personaltax_yearadopted) %>%
  mutate(num_policy_years = sum(personaltax_yearadopted, na.rm = TRUE)) %>%
  mutate(tax_policy = ifelse(is.na(personaltax_yearadopted), FALSE, TRUE)) %>%
  mutate(region = state.region[match(state, state.name)]) %>%
  mutate(state_abb = state.abb[match(state, state.name)]) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = year, y = reorder(state_abb, num_policy_years))) +
  geom_tile(aes(fill = tax_policy),
            color = "white") +
  facet_grid(region ~ ., scales = "free_y", space = "free_y") +
  theme_minimal() +
  theme(legend.justification = "top") +
  scale_fill_manual(values = c('gray90', 'goldenrod1'),
                    name = "Policy in\neffect?") +
  labs(title = "Availability of renewable tax credits varies by state",
       subtitle = "Adoption of renewable tax credit policy by state and year",
       x = "Year",
       y = "State",
       caption = "Data source: IPPSR, Michigan State University")

ggsave(here('output', 'output-intermediate', 'policy_by_state_year.pdf'),
       width = 6, height = 8)
```



## Number of states with personal renewable tax credits by year

```{r tax-plot-2}
# Number of states with personal renewable tax credits by year
policy %>%
  group_by(year) %>%
  summarise(count_with_credits = sum(personaltax_yearadopted, na.rm = TRUE)) %>%
  filter(count_with_credits > 0) %>%
  ggplot() +
  geom_line(aes(x = year, y = count_with_credits, group = 1)) +
  theme_minimal() +
  labs(title = 'What happened in 2001?',
       subtitle = 'Why did 3 states *drop* the policy?',
       x = 'Year',
       y = 'Number of States')
```


```{r tax-plot-3}
# States by year of first adoption of personal tax credits for renewable energy
# Merge on 2016 panels per 10k population
policy_by_start_year <- policy %>%
  select(state, year, personaltax_yearadopted) %>%
  mutate(tax_start_year = ifelse(personaltax_yearadopted == 1, year, NA)) %>%
  group_by(state) %>%
  mutate(tax_start_year = min(tax_start_year, na.rm = TRUE)) %>%
  mutate(tax_start_year = ifelse(is.infinite(tax_start_year), NA, tax_start_year)) %>%
  select(state, tax_start_year) %>%
  unique() %>%
  ungroup() %>%
  mutate(years_since = 2018 - tax_start_year) %>%
  mutate(state = state.abb[match(state, state.name)]) %>%
  left_join(panels_cumul %>% filter(year == 2016) %>% select(state, panels_per_10k),
            by = c("state" = "state"))

# Scatterplot using geom_miss_point - how do I label the missing points?
policy_by_start_year %>%
  ggplot(aes(x = years_since, y = panels_per_10k)) +
  geom_miss_point() +
  geom_text_repel(aes(label = state),
                  na.rm = TRUE) +
  theme_minimal()

# Scatterplot using missing flag, imputed -5 for missing years, geom_text
# policy_by_start_year %>%
#   mutate(no_tax_credits = ifelse(is.na(years_since), TRUE, FALSE)) %>%
#   mutate(years_since = ifelse(is.na(years_since), 0, years_since)) %>%
#   mutate(label = ifelse((panels_per_10k > 50 | years_since > 20), state, "")) %>%
#   ggplot(aes(x = years_since, y = panels_per_10k)) +
#   geom_point(aes(color = no_tax_credits)) +
#   geom_text_repel(aes(label = label)) +
#   theme_minimal()
```

```{r tax-plot-4}
# Bubble chart of panels vs avg DNI, grouped by availability of tax credits,
# sized by population 

panels_cumul %>%
  filter(year == 2016) %>%
  select(state, panels_per_10k) %>%
  inner_join(dni_df %>% 
               filter(period == "Ann") %>%
               mutate(state = state.abb[match(State, state.name)]) %>%
               select(state, avg),
             by = c("state" = "state")) %>%
  inner_join(policy %>%
               filter(year == 2016) %>%
               select(state, personaltax_yearadopted) %>%
               mutate(state = state.abb[match(state, state.name)]) %>%
               mutate(personaltax = ifelse(is.na(personaltax_yearadopted), 
                                                       FALSE, TRUE)),
             by = c("state" = "state")) %>%
  inner_join(pop_by_state %>%
               filter(year == 2016),
             by = c("state", "state")) %>%
  mutate(label = ifelse((panels_per_10k > 50 | avg > 6), state, "")) %>%
  ggplot(aes(x = avg, y = panels_per_10k)) +
  geom_point(aes(color = personaltax, size = total_pop),
             na.rm = TRUE) +
  geom_text_repel(aes(label = label),
                  na.rm = TRUE) +
  scale_color_discrete(name = "Renewable\ntax credits\navailable?") +
  scale_size_continuous(name = "Population\nsize",
                        labels = scales::number) +
  labs(title = "No clear association between solar panel installs and renewable tax credits",
       subtitle = "Cumulative solar panels per 10K people vs. total solar irradiance vs. renewable tax credits (as of 2016)",
       x = "Average Annual DNI (kWh/m2/day)",
       y = "Solar Panels per 10,000 people",
       caption = "Data sources:\nNREL (US Dept of Energy), IPPSR (Michigan State University), National Cancer Institute") +
  theme_minimal()

ggsave(here('output', 'output-intermediate', 'panels_vs_dni_vs_tax.pdf'),
       width = 8, height = 6)
```


# Next steps

-   What is the composition of renewable energy generation? What is the distribution between residential, industrial, corporate sources?
-   What explains why MA, CT, NJ, DE have such high solar panel installation rates despite getting little to no sun?










